---
title: "R Programming Practice Assignment"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Part 1: Conditional Probability with Data Frames and Tables

## Question 1: Survey Data Analysis

### a) Create a data frame with columns: StudentID, Gender, Subject, GradeLevel.
```{r}
# Your R code here

```

### b) Convert it to a contingency table showing Gender vs Subject using table().
```{r}
# Your R code here

```

### c) Use prop.table() to calculate conditional probability P(Math | Male).
```{r}
# Your R code here

```

### d) Calculate the conditional probability P(Female | Science).
```{r}
# Your R code here

```

### e) Find P(Arts | GradeLevel = College).
```{r}
# Your R code here

```

## Question 2: Medical Testing Scenario

### a) Create a data frame with columns: PatientID, Disease (Present/Absent), TestResult (Positive/Negative).
```{r}
# Your R code here

```

### b) Create a contingency table for Disease vs TestResult.
```{r}
# Your R code here

```

### c) Use prop.table() to calculate the sensitivity P(Positive Test | Disease Present).
```{r}
# Your R code here

```

### d) Calculate the specificity P(Negative Test | Disease Absent).
```{r}
# Your R code here

```

### e) Find the positive predictive value P(Disease Present | Positive Test).
```{r}
# Your R code here

```

## Question 3: Customer Purchase Behavior

### a) Create a data frame with columns: CustomerID, AgeGroup (Young/Middle/Senior), PurchaseCategory (Electronics/Clothing/Books).
```{r}
# Your R code here

```

### b) Make a contingency table for AgeGroup vs PurchaseCategory.
```{r}
# Your R code here

```

### c) Calculate P(Electronics | Young) using prop.table().
```{r}
# Your R code here

```

### d) Find P(Young | Electronics).
```{r}
# Your R code here

```

### e) Compare purchase patterns across age groups using proportions.
```{r}
# Your R code here

```


# Part 2: Data Visualization

## Question 4: Sales Performance Analysis

### a) Create a data frame with columns: Rep (Alice, Bob, Carol), Quarter (Q1–Q4), Sales.
```{r}
# Your R code here

```

### b) Make a bar plot showing total sales by representative using barplot().
```{r}
# Your R code here

```

### c) Create a bar plot showing quarterly sales for each rep.
```{r}
# Your R code here

```

### d) Add appropriate titles, labels, and colors to your plots.
```{r}
# Your R code here

```

### e) Make a horizontal bar plot of average quarterly sales by rep.
```{r}
# Your R code here

```

## Question 5: Student Grade Distribution

### a) Create a data frame with columns: StudentID, Subject (Math/Science/English), Grade.
```{r}
# Your R code here

```

### b) Make histograms for each subject's grade distribution using hist().
```{r}
# Your R code here

```

### c) Create box plots comparing grade distributions across subjects using boxplot().
```{r}
# Your R code here

```

### d) Make a bar plot showing average grades by subject.
```{r}
# Your R code here

```

### e) Add proper titles, labels, and colors to all visualizations.
```{r}
# Your R code here

```

## Question 6: Temperature Data Analysis

### a) Create a data frame with columns: Day, Season (Spring/Summer/Fall/Winter), Temperature.
```{r}
# Your R code here

```

### b) Create histograms for temperature in each season.
```{r}
# Your R code here

```

### c) Make box plots showing temperature distribution by season.
```{r}
# Your R code here

```

### d) Create a bar plot showing average temperature by season.
```{r}
# Your R code here

```

### e) Use different colors for each season and add proper labels.
```{r}
# Your R code here

```


# Part 3: Integration Challenges

## Question 7: Store Sales Analysis

### a) Create a data frame with columns: Month, Product (Electronics/Clothing/Books), Sales, Customers.
```{r}
# Your R code here

```

### b) Calculate conditional probabilities: P(High Sales | Electronics), P(Electronics | Month = January).
```{r}
# Your R code here

```

### c) Create bar plots comparing sales by product and by month.
```{r}
# Your R code here

```

### d) Make box plots showing sales distribution by product category.
```{r}
# Your R code here

```

### e) Write a brief interpretation of your findings.
```{r}
# Your R code here

```

## Question 8: Survey Data Integration

### a) Create a data frame with 40 respondents: Age (Young/Old), Product (A/B), Satisfaction (High/Low).
```{r}
# Your R code here

```

### b) Calculate P(High Satisfaction | Product A), P(Product A | Young).
```{r}
# Your R code here

```

### c) Create contingency tables and use prop.table() for analysis.
```{r}
# Your R code here

```

### d) Make bar plots and box plots to visualize relationships.
```{r}
# Your R code here

```

### e) Summarize your findings about customer satisfaction patterns.
```{r}
# Your R code here

```


# Part 4: Final Challenge

## Question 9: Complete Data Analysis Project

### a) Create a data frame with student performance data (30 students, 2 subjects, with gender and study method).
```{r}
# Your R code here

```

### b) Perform conditional probability analysis using table() and prop.table().
```{r}
# Your R code here

```

### c) Create multiple types of visualizations (bar plots, histograms, box plots).
```{r}
# Your R code here

```

### d) Calculate and interpret at least 3 different conditional probabilities.
```{r}
# Your R code here

```

### e) Write a summary of your findings with supporting visualizations.
```{r}
# Your R code here

```
